Speeding up decimal multiplication

### 2020-11-18 paper/paper.tex
Update

We now define the cyclic convolution.
Consider an aribtrary algebraic ring $\mathcal{R}.$
The cyclic convolution $x \star y, \, (\_ \star \_) \colon \mathcal{R}^n \times \mathcal{R}^n \to \mathcal{R}^n$ is defined as follows:
$$(x \star y)[k] = \sum\limits_{i=0}^{n - 1} x[i] y[(k - i) \mymod n].$$

@@

Note that $c_i \le (B-1)^2 \min\{n, m\}$ as $x_i < B, \, y_i < B,$
and, because of our zero padding, the summation in the definition of the cyclic convolution
can have at most $\min\{n,m\}$ non-zero summands.

The representation of $z$ in base $B$ can then be recovered by the following iterative process:
\begin{eqnarray*}
\sigma_0 &=& 0; \\
z_i &=& (\sigma_i + c_i) \mymod B; \\
\sigma_{i+1} &=& (\sigma_i + c_i) \mydiv B.
\end{eqnarray*}

@@

Define $L = (B-1)^2 \min\{n,m\}$ and remember $c_i \le L.$
Define also the following sequence:
$$a_0 = 0, \, a_{i+1} = (a_i + L) \mydiv B.$$

@@

We have
$$\sigma_i \le a_i \le a^\prime_i < \lim\limits_{i \to \infty} a'_i = \frac{L}{B - 1} = (B-1)\min\{n, m\}.$$

@@

Discrete Fourier transform $\mathfrak{f} \colon \mathcal{F}^N \to \mathcal{F}^N$ is then defined as follows:
$$\mathfrak{f}(x)[k] = \sum\limits_{i=0}^{N-1} x[i] \xi^{ik}.$$

Inverse discrete Fourier transform $\mathfrak{f}^{-1} \colon \mathcal{F}^N \to \mathcal{F}^N$ is defined similarly,
up to the negated exponent of $\xi$ and multiplication by $N^{-1}$:
$$\mathfrak{f}^{-1}(x)[k] = N^{-1} \sum\limits_{i=0}^{N-1} x[i] \xi^{-ik}.$$

It is easy to see that both $\mathfrak{f}$ and $\mathfrak{f}^{-1}$ are linear, which means that for any
$\alpha \in \mathcal{F}$ and $x \in \mathcal{F}^N,$
$\mathfrak{f}(\alpha x) = \alpha \mathfrak{f}(x)$
and
$\mathfrak{f}^{-1}(\alpha x) = \alpha \mathfrak{f}^{-1}(x).$ Here, the product of a scalar and a vector
$\alpha v = \alpha \langle v_0, \ldots, v_{n-1} \rangle$ denotes, as usual,
$\langle \alpha v_0, \ldots, \alpha v_{n-1} \rangle.$

The convolution theorem says that, for any two vectors $x, y \in \mathcal{F}^N,$
$$x \star y = \mathfrak{f}^{-1}\big( \mathfrak{f}(x) \cdot \mathfrak{f}(y) \big),$$
where $\cdot$ denotes scalar (element-wise) product.

Note that $\star$ is bilinear, which means that it is linear in both of its arguments:
for any scalar $\alpha \in \mathcal{F}$ and any two vectors $x, y \in \mathcal{F}^N,$
$$x \star (\alpha y) = (\alpha x) \star y = \alpha (x \star y).$$

@@

Another approach is to pick $n$ pairwise distinct primes, $p_1, \ldots, p_n,$ such that
$$\prod\limits_{i=1}^{n} p_i > (B-1)^2 M.$$
Then, for each $i,$ compute the convolution vector in $\mathrm{F}_{p_i}$ and use
the Chinese remainder theorem to recover the actual answer.
In this case, we would need to round the size of the convolution, $N,$ up to some divisor of
$$\gcd(p_1 - 1, \ldots, p_n - 1).$$
We thus need to choose $p_1, \ldots, p_n$ such that this value is highly composite.

@@

Let us now choose the number of primes, $\ell.$
Remember $\mu = 2^w,$ where $w$ is the length of the machine word in bits.
Remember also that we want to pick $p_1, \ldots, p_\ell$ such that
$$\prod\limits_{i=1}^{\ell} p_i > (B-1)^2 M,$$
where $M$ is the maximum length of a multiplicand possible.

We will require $p_i < \mu/2$ for simpler implementation of addition modulo $p_i.$
If this does not hold, then, during addition of two numbers modulo $p_i,$ the raw sum may overflow
the machine word, and the implementation would need to check for two conditions instead of just
one: we would need to subtract $p_i$ (modulo $\mu$) from the result if either the overflow happened
or the result is greater or equal to $p_i.$

Having fixed $\mu = 2^w$ and $M,$ and assuming all $p_1, \ldots, p_\ell$ will be
approximately equal to $\mu/2,$ we can define the following function:
$$\lambda(\ell) = \max\{n \in \mathbb{N} : ({10}^n - 1)^2 < (\mu / 2)^\ell / M\}.$$
Then, once we choose $\ell,$ we calculate $\lambda(\ell)$ and, assuming it is positive, we can put
$B = {10}^{\lambda(\ell)}.$ If $\lambda(\ell) = 0,$ the chosen value of $\ell$ is too small; more
prime moduli are needed.

@@

The definition of $\lambda(\ell)$ in the previous subsection gives us a way to calculate
$B = 10^{\lambda(\ell)}$ given the values of $\mu,$ $M,$ and $\ell.$
In practice, it is better to support multiple bases, each for its own range of transform lengths.
This eliminates the need for picking the single maximum transform length $M$ and slowing down
smaller transforms.

@@

Remember we defined $w$ as the bit width of the machine word;
define then $R = 2^w \in \mathrm{F}_p.$ We assume $p > 2,$ so $R$ is non-zero in $\mathrm{F}_p.$

@@

The \textit{Montgomery reduction} is a function
$\redc \colon \mathrm{F}_p \times \mathrm{F}_p \to \mathrm{F}_p$
defined as follows:
$$\redc(x, y) = R^{-1} x y.$$
It is important because the Montgomery representation $R(x y)$ of the product $x y$ is
$\redc(Rx, Ry).$ Also, any $x \in \mathrm{F}_p$ can be converted into Montgomery
representation by invoking $\redc(x, R^2) = Rx;$ and out of Montgomery
representation by invoking $\redc(x, 1) = R^{-1}x.$

@@

\begin{procedure}[H]
\DontPrintSemicolon
\caption{MontgomeryReduce($a$)}
\KwData{
    Integer $\mu.$\;
    Integer $p$ coprime with $\mu.$\;
    Integer $p' \in [0; \mu-1]$ such that $p \cdot p' \equiv -1 \, (\mymod \mu).$
}
\KwIn{
    Integer $a \in [0; \mu p - 1].$
}
\KwOut{
    Integer $b \in [0; p-1]$ such that $b \equiv a \mu^{-1} \, (\mymod p).$
}
\Begin{
    $m \longleftarrow ((a \mymod \mu) \cdot p') \mymod \mu$\;
    $r \longleftarrow (a + m \cdot p) \mydiv \mu$
        \tcc*[r]{this division is exact} % apparently no need to use \; after \tcc
    \eIf{$r \ge p$}{
        \Return{$r - p$}\;
    }{
        \Return{$r$}\;
    }
}
\end{procedure}

The value of $p'$ can be pre-calculated with extended Euclidean algorithm for GCD;
we have $\gcd(p, \mu) = 1,$ so $\exists A, B \in \mathbb{Z} \colon A p + B \mu = 1.$
This means $A p \equiv 1 \, (\mymod \mu)$ and $-Ap \equiv -1 \, (\mymod \mu),$ so
$$p' = (-Ap) \mymod \mu.$$

@@

Having chosen the Montgomery reduction, we are going search for primes of form
$$c \cdot 3 \cdot 2^n + 1.$$

@@

Observe that, in order to guarantee the uniqueness of representation of a prime as
$c \cdot 3 \cdot 2^n + 1,$
we need to require $c$ to be odd --- otherwise a power of two can be factored out into $2^n.$

Define $n_{\mathrm{min}} = \lceil \log_2(M/3) \rceil,$ the lower bound for $n$ in the formula for
the desired form of our primes above.
We have, then, the following requirements for each $p_i$:
\begin{enumerate}
    \item $p_i$ is of form $c \cdot 3 \cdot 2^n + 1,$ where $c$ is odd and $n \ge n_{\mathrm{min}};$
    \item $p_i < \mu / 2.$
\end{enumerate}
Of all possible primes with those properties, we need to pick the $\ell$ largest.

@@

Both algorithms take two parameters, an array $A$ of field elements and a field element
$\omega = \xi^{\pm 1},$
and return an array $A'$ of length $|A'| = |A|$
such that $$A'[k] = \sum\limits_{i=0}^{|A| - 1} A[i] \omega^{ik}.$$
If we express this transform as a function $\tau(A,\omega),$ then DFT and IDFT can be expressed in
terms of $\tau$ as follows:
\begin{eqnarray*}
\mathfrak{f}(x) &=& \tau(x, \xi); \\
\mathfrak{f}^{-1}(x) &=& N^{-1} \tau(x, \xi^{-1}),
\end{eqnarray*}
where $N$ is the length of the transform.

@@

The four-line transform on $A[k+j]$ and $A[k+j+m/2]$ is known as ``butterfly'', or,
more specifically, this one is the Cooley-Tukey butterfly; it maps
$\langle x,y \rangle \mapsto \langle x', y' \rangle$ as follows:
\begin{eqnarray*}
x' &=& x + \varphi \cdot y; \\
y' &=& x - \varphi \cdot y.
\end{eqnarray*}

@@

The four-line transform on $A[k+j]$ and $A[k+j+m/2]$ is known as the Gentleman-Sande butterfly; it maps
$\langle x,y \rangle \mapsto \langle x', y' \rangle$ as follows:
\begin{eqnarray*}
x' &=& x + y; \\
y' &=& \varphi \cdot (x - y).
\end{eqnarray*}

@@

Remember we expressed the cyclic convolution in terms of DFT and IDFT as follows:
$$x \star y = \mathfrak{f}^{-1}\big( \mathfrak{f}(x) \cdot \mathfrak{f}(y) \big),$$
where $\cdot$ denotes scalar (element-wise) product.

Note that, for the computation of $(\_ \star \_),$ the specific order in which the direct transform
$\mathfrak{f}$ produces its outputs, and the order in which the inverse transform
$\mathfrak{f}^{-1}$ expects its inputs to be in, do not matter; we only need these two orders
to match.
Indeed, $(\_ \cdot \_)$ is element-wise, $(x \cdot y)[i] = x[i] y[i],$
so, for any permutation $\pi$ we have $$\pi^{-1}(\pi(x) \cdot \pi(y)) = x \cdot y.$$

@@

Let us write $T_s$ for all $s$ into a single array $T_\ast$
of length $2^N-1 = |A|-1$ so that
$$T_s[i] = T_\ast[(2^0 + 2^1 + \cdots + 2^{s-2}) + i] = T_\ast[2^{s-1} - 1 + i].$$
Then, given $j$ and $m=2^s,$ the factor $\varphi$ can be expressed as
$T_\ast[\Delta_s + j],$ where $\Delta_s = 2^{s-1}-1 = m/2 - 1.$

@@

We are now returning to the Montgomery reduction.

Remember we defined $R = 2^w \in \mathrm{F}_p,$ where $w$ is the length of machine word in bits;
we assume $p > 2,$ so $R$ is non-zero in $\mathrm{F}_p.$

Remember also we defined the function
$\redc \colon \mathrm{F}_p \times \mathrm{F}_p \to \mathrm{F}_p$
in the following way:
$$\redc(x, y) = R^{-1} x y.$$

@@

We can consider, then,
$\widetilde{\mathfrak{f}}$ and $\widetilde{\mathfrak{f}}^{-1},$
the DFT and IDFT, correspondingly, for the field $\widetilde{\mathrm{F}}_p.$
We have the following identities, by the isomorphism, for any $x \in \mathrm{F}_p^N$:
\begin{eqnarray*}
    \mathfrak{f}(x)         &=& R^{-1} \, \widetilde{\mathfrak{f}}(R \, x); \\
    \mathfrak{f}^{-1}(x)    &=& R^{-1} \, \widetilde{\mathfrak{f}}^{-1}(R \, x).
\end{eqnarray*}

Rewrite with $x = R^{-1}y$:
\begin{eqnarray*}
    \mathfrak{f}(R^{-1}y)       &=& R^{-1} \, \widetilde{\mathfrak{f}}(y); \\
    \mathfrak{f}^{-1}(R^{-1}y)  &=& R^{-1} \, \widetilde{\mathfrak{f}}^{-1}(y).
\end{eqnarray*}

Rewrite the left-hand sides by the linearity of $\mathfrak{f}$ and $\mathfrak{f}^{-1}$:
\begin{eqnarray*}
    R^{-1} \mathfrak{f}(y)       &=& R^{-1} \, \widetilde{\mathfrak{f}}(y); \\
    R^{-1} \mathfrak{f}^{-1}(y)  &=& R^{-1} \, \widetilde{\mathfrak{f}}^{-1}(y).
\end{eqnarray*}

Dividing both sides by $R^{-1},$ we see that $\mathfrak{f}(y)$ and $\widetilde{\mathfrak{f}}(y)$
coincide for any $y = Rx;$ and so do $\mathfrak{f}^{-1}(y)$ and $\widetilde{\mathfrak{f}}^{-1}(y).$

Let us now consider the cyclic convolution $\widetilde{\star}$ for $\widetilde{\mathrm{F}}_p;$
we know that
$$x \, \widetilde{\star} \, y = \widetilde{\mathfrak{f}}^{-1}
\big( \thinspace \widetilde{\mathfrak{f}}(x) \, \widetilde{\otimes} \, \widetilde{\mathfrak{f}}(y) \big),$$
where $\widetilde{\otimes}$ denotes element-wise $\redc$
(which is, remember, the multiplication operation of $\widetilde{\mathrm{F}}_p.$)
Having proven that
$\widetilde{\mathfrak{f}} \equiv \mathfrak{f}$
and
$\widetilde{\mathfrak{f}}^{-1} \equiv \mathfrak{f}^{-1}$
above, we can rewrite it as follows:
$$x \, \widetilde{\star} \, y = R^{-1} \mathfrak{f}^{-1}\big( \mathfrak{f}(x) \cdot \mathfrak{f}(y) \big),$$
where $\cdot$ denotes element-wise product.
This, in part, can be rewritten as
$$x \, \widetilde{\star} \, y = R^{-1}(x \star y).$$

@@

Remember in section~\ref{section:omitting1} we established that, with
$x \cdot y$ denoting scalar (element-wise) product between $x,y \in \mathcal{F}^N,$
for any permutation $\pi$ we have $$\pi^{-1}(\pi(x) \cdot \pi(y)) = x \cdot y.$$

We then used this property to get rid of bit-reversal permutations in DIT and DIF.
It turns out that, for both four-step and six-step algorithms,
the same property can be employed to get rid of the
final transposition step in DFT and the initial transposition step in IDFT simultaneously.

@@

@@

\subsection{Power table}

>>>>

We now define the cyclic convolution.
Consider an aribtrary algebraic ring $\mathcal{R}.$
The cyclic convolution $x \star y, \, (\_ \star \_) \colon \mathcal{R}^n \times \mathcal{R}^n \to \mathcal{R}^n$ is defined as follows:
\begin{equation} \label{eq:star}
(x \star y)[k] = \sum\limits_{i=0}^{n - 1} x[i] y[(k - i) \mymod n].
\end{equation}

@@

Note that
\begin{equation} \label{eq:ci_bound}
c_i \le (B-1)^2 \min\{n, m\}
\end{equation}
as both $x_i$ and $y_i$ are less that $B$ and, because of our zero padding, the summation
in~\eqref{eq:star} can have at most $\min\{n,m\}$ non-zero summands.

The representation of $z$ in base $B$ can then be recovered by the following iterative process:
\begin{subequations} \label{eq:sigmai_and_zi}
\begin{align}
\sigma_0 &= 0; \\
z_i &= (\sigma_i + c_i) \mymod B; \\
\sigma_{i+1} &= (\sigma_i + c_i) \mydiv B.
\end{align}
\end{subequations}

@@

Define $L = (B-1)^2 \min\{n,m\}$ and remember that in~\eqref{eq:ci_bound}, we established that
$c_i \le L.$
Define also the following sequence:
$$a_0 = 0, \, a_{i+1} = (a_i + L) \mydiv B.$$

@@

We have
\begin{equation} \label{eq:sigmai_bound}
\sigma_i \le a_i \le a^\prime_i < \lim\limits_{i \to \infty} a'_i = \frac{L}{B - 1} = (B-1)\min\{n, m\}.
\end{equation}

@@

Discrete Fourier transform $\mathfrak{f} \colon \mathcal{F}^N \to \mathcal{F}^N$ is then defined as follows:
\begin{equation} \label{eq:dft}
\mathfrak{f}(x)[k] = \sum\limits_{i=0}^{N-1} x[i] \xi^{ik}.
\end{equation}

Inverse discrete Fourier transform $\mathfrak{f}^{-1} \colon \mathcal{F}^N \to \mathcal{F}^N$ is defined similarly,
up to the negated exponent of $\xi$ and multiplication by $N^{-1}$:
\begin{equation} \label{eq:idft}
\mathfrak{f}^{-1}(x)[k] = N^{-1} \sum\limits_{i=0}^{N-1} x[i] \xi^{-ik}.
\end{equation}

It is easy to see that both $\mathfrak{f}$ and $\mathfrak{f}^{-1}$ are linear, which means that for any
$\alpha \in \mathcal{F}$ and $x \in \mathcal{F}^N,$
\begin{subequations} \label{eq:transform_linearity}
\begin{align}
\mathfrak{f}(\alpha x) &= \alpha \mathfrak{f}(x); \\
\mathfrak{f}^{-1}(\alpha x) &= \alpha \mathfrak{f}^{-1}(x).
\end{align}
\end{subequations}
Above, the product of a scalar and a vector
$\alpha v = \alpha \langle v_0, \ldots, v_{n-1} \rangle$ denotes, as usual,
$\langle \alpha v_0, \ldots, \alpha v_{n-1} \rangle.$

The convolution theorem says that, for any two vectors $x, y \in \mathcal{F}^N,$
\begin{equation} \label{eq:conv_theorem}
x \star y = \mathfrak{f}^{-1}\big( \mathfrak{f}(x) \cdot \mathfrak{f}(y) \big),
\end{equation}
where $\cdot$ denotes scalar (element-wise) product.

Note that $\star$ is bilinear, which means that it is linear in both of its arguments:
for any scalar $\alpha \in \mathcal{F}$ and any two vectors $x, y \in \mathcal{F}^N,$
\begin{equation} \label{eq:star_linearity}
x \star (\alpha y) = (\alpha x) \star y = \alpha (x \star y).
\end{equation}

@@

Another approach is to pick $n$ pairwise distinct primes, $p_1, \ldots, p_n,$ such that
\begin{equation} \label{eq:product_inequality}
\prod\limits_{i=1}^{n} p_i > (B-1)^2 M.
\end{equation}
Then, for each $i,$ compute the convolution vector in $\mathrm{F}_{p_i}$ and use
the Chinese remainder theorem to recover the actual answer.
In this case, we would need to round the size of the convolution, $N,$ up to some divisor of
$$\gcd(p_1 - 1, \ldots, p_n - 1).$$
We thus need to choose $p_1, \ldots, p_n$ such that this value is highly composite.

@@

Let us now choose the number of primes, $\ell.$
Remember we defined $\mu = 2^w,$ where $w$ is the length of the machine word in bits.
Remember also that by~\eqref{eq:product_inequality},
we want to pick $p_1, \ldots, p_\ell$ such that
$$\prod\limits_{i=1}^{\ell} p_i > (B-1)^2 M,$$
where $M$ is the maximum length of a multiplicand possible.

We will require
\begin{equation} \label{eq:pi_bound}
p_i < \mu/2
\end{equation}
for simpler implementation of addition modulo $p_i.$
If this does not hold, then, during addition of two numbers modulo $p_i,$ the raw sum may overflow
the machine word, and the implementation would need to check for two conditions instead of just
one: we would need to subtract $p_i$ (modulo $\mu$) from the result if either the overflow happened
or the result is greater or equal to $p_i.$

Having fixed $\mu = 2^w$ and $M,$ and assuming all $p_1, \ldots, p_\ell$ will be
approximately equal to $\mu/2,$ we can define the following function:
\begin{equation} \label{eq:lambda}
\lambda(\ell) = \max\{n \in \mathbb{N} : ({10}^n - 1)^2 < (\mu / 2)^\ell / M\}.
\end{equation}
Then, once we choose $\ell,$ we calculate $\lambda(\ell)$ and, assuming it is positive, we can put
$B = {10}^{\lambda(\ell)}.$ If $\lambda(\ell) = 0,$ the chosen value of $\ell$ is too small; more
prime moduli are needed.

@@

Definition~\eqref{eq:lambda} gives us a way to calculate
$B = 10^{\lambda(\ell)}$ given the values of $\mu,$ $M,$ and $\ell.$
In practice, it is better to support multiple bases, each for its own range of transform lengths.
This eliminates the need for picking the single maximum transform length $M$ and slowing down
smaller transforms.

@@

Remember we defined $w$ as the bit width of the machine word;
define then
\begin{equation} \label{eq:mont_r}
R = 2^w \in \mathrm{F}_p.
\end{equation}
We assume $p > 2,$ so $R$ is non-zero in $\mathrm{F}_p.$

@@

The \textit{Montgomery reduction} is a function
$\redc \colon \mathrm{F}_p \times \mathrm{F}_p \to \mathrm{F}_p$
defined as follows:
\begin{equation} \label{eq:mont_redc}
\redc(x, y) = R^{-1} x y.
\end{equation}
It is important because the Montgomery representation $R(x y)$ of the product $x y$ is
$\redc(Rx, Ry).$ Also, any $x \in \mathrm{F}_p$ can be converted into Montgomery
representation by invoking $\redc(x, R^2) = Rx;$ and out of Montgomery
representation by invoking $\redc(x, 1) = R^{-1}x.$

@@

\begin{procedure}[H]
\DontPrintSemicolon
\caption{MontgomeryReduce($a$)}
\KwData{
    Integer $\mu.$\;
    Integer $p$ coprime with $\mu.$\;
    Integer $p' \in [0; \mu-1]$ such that $p \cdot p' \equiv -1 \, (\mymod \, \mu).$
}
\KwIn{
    Integer $a \in [0; \mu p - 1].$
}
\KwOut{
    Integer $b \in [0; p-1]$ such that $b \equiv a \mu^{-1} \, (\mymod \, p).$
}
\Begin{
    $m \longleftarrow ((a \mymod \mu) \cdot p') \mymod \mu$\;
    $r \longleftarrow (a + m \cdot p) \mydiv \mu$
        \tcc*[r]{this division is exact} % apparently no need to use \; after \tcc
    \eIf{$r \ge p$}{
        \Return{$r - p$}\;
    }{
        \Return{$r$}\;
    }
}
\end{procedure}

For the pre-calculated data, put $\mu = 2^w,$ $p > 2$ is our prime number, and
the value of $p'$ can be calculated with the extended Euclidean algorithm for GCD;
we have $\gcd(p, \mu) = 1,$ so $\exists c_1, c_2 \in \mathbb{Z} \colon c_1 p + c_2 \mu = 1.$
This means $c_1 p \equiv 1 \, (\mymod \, \mu)$ and $-c_1 p \equiv -1 \, (\mymod \, \mu),$ so
$$p' = (-c_1 p) \mymod \mu.$$

@@

Having chosen the Montgomery reduction, we are going search for primes of form
\begin{equation} \label{eq:pformula}
c \cdot 3 \cdot 2^n + 1.
\end{equation}

@@

Observe that, in order to guarantee the uniqueness of representation of a prime as~\eqref{eq:pformula},
we need to require $c$ to be odd --- otherwise a power of two can be factored out into $2^n.$

Define $n_{\mathrm{min}} = \lceil \log_2(M/3) \rceil,$ the lower bound for $n$ in~\eqref{eq:pformula}.
We have, then, the following requirements for each $p_i$:
\begin{enumerate}
    \item $p_i$ is of form $c \cdot 3 \cdot 2^n + 1,$ where $c$ is odd and $n \ge n_{\mathrm{min}};$
    \item $p_i < \mu / 2.$
\end{enumerate}
Of all possible primes with those properties, we need to pick the $\ell$ largest.

@@

Both algorithms take two parameters, an array $A$ of field elements and a field element
$\omega = \xi^{\pm 1},$
and return an array $A'$ of length $|A'| = |A|$
such that $$A'[k] = \sum\limits_{i=0}^{|A| - 1} A[i] \omega^{ik}.$$
If we express this transform as a function $\tau(A,\omega),$ then DFT and IDFT of length $N$ can be
expressed in terms of $\tau$ as follows:
\begin{subequations}
\begin{align}
\mathfrak{f}(x) &= \tau(x, \xi); \\
\mathfrak{f}^{-1}(x) &= N^{-1} \tau(x, \xi^{-1}).
\end{align}
\end{subequations}

@@

The four-line transform on $A[k+j]$ and $A[k+j+m/2]$ is known as ``butterfly'', or,
more specifically, this one is the Cooley-Tukey butterfly; it maps
$\langle x,y \rangle \mapsto \langle x', y' \rangle$ as follows:
\begin{subequations}
\begin{align}
x' &= x + \varphi \cdot y; \\
y' &= x - \varphi \cdot y.
\end{align}
\end{subequations}

@@

The four-line transform on $A[k+j]$ and $A[k+j+m/2]$ is known as the Gentleman-Sande butterfly; it maps
$\langle x,y \rangle \mapsto \langle x', y' \rangle$ as follows:
\begin{subequations}
\begin{align}
x' &= x + y; \\
y' &= \varphi \cdot (x - y).
\end{align}
\end{subequations}

@@

Note that in~\eqref{eq:conv_theorem}, for the computation of $(\_ \star \_),$ the specific order in
which the direct transform $\mathfrak{f}$ produces its outputs, and the order in which the inverse
transform $\mathfrak{f}^{-1}$ expects its inputs to be in, do not matter; we only need these two
orders to match.
Indeed, $(\_ \cdot \_)$ is element-wise, $(x \cdot y)[i] = x[i] y[i],$
so, for any permutation $\pi$ we have
\begin{equation} \label{eq:cdot_permutation_invariance}
\pi^{-1}(\pi(x) \cdot \pi(y)) = x \cdot y.
\end{equation}

@@

Let us write $T_s$ for all $s$ into a single array $T_\ast$
of length $2^N-1 = |A|-1$ so that
$$T_s[i] = T_\ast[(2^0 + 2^1 + \cdots + 2^{s-2}) + i] = T_\ast[2^{s-1} - 1 + i].$$
Then, given $j$ and $m=2^s,$ the factor $\varphi$ can be expressed as
$T_\ast[\delta + j],$ where $\delta = 2^{s-1}-1 = m/2 - 1.$

@@

We are now returning to the Montgomery reduction; we defined $R$ and $\redc$
in~\ref{section:montgomery_intro}; for the purposes of this section, $p > 2$ is arbitrary prime.

@@

We can consider, then,
$\widetilde{\mathfrak{f}}$ and $\widetilde{\mathfrak{f}}^{-1},$
the DFT and IDFT, correspondingly, for the field $\widetilde{\mathrm{F}}_p.$
We have the following identities, by the isomorphism, for any $x \in \mathrm{F}_p^N$:
\begin{subequations}
\begin{align}
    \mathfrak{f}(x)         &= R^{-1} \, \widetilde{\mathfrak{f}}(R \, x); \\
    \mathfrak{f}^{-1}(x)    &= R^{-1} \, \widetilde{\mathfrak{f}}^{-1}(R \, x).
\end{align}
\end{subequations}

Rewrite with $x = R^{-1}y$:
\begin{subequations}
\begin{align}
    \mathfrak{f}(R^{-1}y)       &= R^{-1} \, \widetilde{\mathfrak{f}}(y); \\
    \mathfrak{f}^{-1}(R^{-1}y)  &= R^{-1} \, \widetilde{\mathfrak{f}}^{-1}(y).
\end{align}
\end{subequations}

Rewrite the left-hand sides by the linearity of $\mathfrak{f}$ and $\mathfrak{f}^{-1}$
(see~\eqref{eq:transform_linearity}):
\begin{subequations}
\begin{align}
    R^{-1} \mathfrak{f}(y)       &= R^{-1} \, \widetilde{\mathfrak{f}}(y); \\
    R^{-1} \mathfrak{f}^{-1}(y)  &= R^{-1} \, \widetilde{\mathfrak{f}}^{-1}(y).
\end{align}
\end{subequations}

Dividing both sides by $R^{-1},$ we see that the transforms in $\mathrm{F}_p$ and
$\widetilde{\mathrm{F}}_p$ completely coincide: for any
$y \in \mathrm{F}_p^N,$
\begin{subequations} \label{eq:dft_idft_in_tildefp}
\begin{align}
\mathfrak{f}(y) &= \widetilde{\mathfrak{f}}(y); \\
\mathfrak{f}^{-1}(y) &= \widetilde{\mathfrak{f}}^{-1}(y).
\end{align}
\end{subequations}

Let us now consider the cyclic convolution $\widetilde{\star}$ for $\widetilde{\mathrm{F}}_p;$
by~\eqref{eq:conv_theorem}, we know that
$$x \, \widetilde{\star} \, y = \widetilde{\mathfrak{f}}^{-1}
\big( \thinspace \widetilde{\mathfrak{f}}(x) \, \widetilde{\otimes} \, \widetilde{\mathfrak{f}}(y) \big),$$
where $\widetilde{\otimes}$ denotes element-wise $\redc$
(which is, remember, the multiplication operation of $\widetilde{\mathrm{F}}_p.$)
Having proven~\eqref{eq:dft_idft_in_tildefp}, we can rewrite it as follows:
$$x \, \widetilde{\star} \, y = R^{-1} \mathfrak{f}^{-1}\big( \mathfrak{f}(x) \cdot \mathfrak{f}(y) \big),$$
where $\cdot$ denotes element-wise product.
This, in part, can be rewritten as
$$x \, \widetilde{\star} \, y = R^{-1}(x \star y).$$

@@

Remember in section~\ref{section:omitting1} we established~\eqref{eq:cdot_permutation_invariance}.
We then used this property to get rid of bit-reversal permutations in DIT and DIF.
It turns out that, for both four-step and six-step algorithms,
the same property can be employed to get rid of the
final transposition step in DFT and the initial transposition step in IDFT simultaneously.

@@

Suppose we want to transpose a matrix with $n$ rows and $2n$ columns in-place.

Let $M$ be the matrix we want to transpose.
Split $M$ into two $n$-by-$n$ submatrices, left submatrix $A$ and right submatrix $B$:
$$
M = \begin{bmatrix}
A & B
\end{bmatrix}.
$$

Note that
$$
M^T =
\begin{bmatrix}
    A & B
\end{bmatrix}^T =
\begin{bmatrix}
    A^T \\
    B^T
\end{bmatrix}.
$$

Let us refer to the rows of $A^T$ as $\alpha_1, \ldots, \alpha_n;$
and to the rows of $B^T$ as $\beta_1, \ldots, \beta_n.$
We write $\longleftrightarrow$ for ``equivalent in flat row-major order to''.
Then,
$$
\begin{bmatrix}
    A^T & B^T
\end{bmatrix} =
\begin{bmatrix}
    \alpha_1 & \beta_1 \\
    \ldots & \ldots \\
    \alpha_n & \beta_n
\end{bmatrix}
\longleftrightarrow
\langle \alpha_1, \beta_1, \alpha_2, \beta_2, \ldots, \alpha_n, \beta_n \rangle.
$$
It differs from
$$
M^T =
\begin{bmatrix}
    A^T \\
    B^T
\end{bmatrix} =
\begin{bmatrix}
    \alpha_1  \\
    \ldots \\
    \alpha_n \\
    \beta_1 \\
    \ldots \\
    \beta_n
\end{bmatrix}
\longleftrightarrow
\langle \alpha_1, \alpha_2, \ldots, \alpha_n, \beta_1, \beta_2, \ldots, \beta_n \rangle.
$$

So we can propose the following algorithm for transposing $M,$ although quite inefficient:
\begin{enumerate}
    \item Interpreting the whole array as $n \times 2n$ matrix, transpose the left and right submatrices.
    \item Interpreting the whole array as $2n \times n$ matrix, permute its \textit{rows} in a certain way.
Namely, we need to apply the permutation $\rho$ that sends a row with zero-based index
$i, 0 \le i < 2n,$ to the zero-based index $$n \cdot (i \mymod 2) + (i \mydiv 2).$$
\end{enumerate}

But permuting \textit{rows} of a matrix means permuting \textit{columns} of the transposed matrix
in the same way. To write it in a formal way, we need to introduce some new notation.

If $\pi$ is a permutation on length-$N$ sequences,
then $\pi' X,$ where $X$ is a matrix with $N$ rows, means $X$ with rows permuted by $\pi;$
$\pi'' Y,$ where $Y$ is a matrix with $N$ columns, means $Y$ with columns permuted by $\pi.$

We can then write formally, for any matrix $X$ with $N$ columns and permutation $\pi$ on length-$N$
sequences,
\begin{equation} \label{eq:matrix_permutation_rule}
(\pi'' X)^T = \pi'(X^T).
\end{equation}

Define the result of transposition of both left and right submatrices of a $n$-by-$2n$ matrix $M$ as
$\Psi(M).$ Note this is step 1 of our algorithm. We have:
$$\rho' \Psi(M) = M^T.$$
Substitute with $M = \rho'' L:$
$$\rho' \Psi(\rho'' L) = (\rho'' L)^T.$$
Apply~\eqref{eq:matrix_permutation_rule} to the right-hand side:
$$\rho' \Psi(\rho'' L) = \rho'(L^T).$$
Simplify, since $\rho'$ is a bijection:
$$\Psi(\rho'' L) = L^T.$$

It means that we can apply $\rho$ to \textit{columns} (as opposed to rows) \textit{before} the transpositions of
submatrices (as opposed to after) --- and still get the transposed matrix as the result.

Application of $\rho$ to a column can be done in a cache-efficient way, and requires $2n$ additional memory.

Since $(M^T)^T = M,$ the transposition of $2n \times n$ matrix is the inverse operation on the array of
the same length; it can be done as doing the inverse operations in reverse order: apply $\rho^{-1}$
to columns, then transpose the submatrices. Application of $\rho^{-1}$ to a column can also be done in a
cache-efficient way, and also requires $2n$ additional memory.

This approach generalizes trivially to the case of $n \times C n$ and $C n \times n$ matrices.
It requires $C n$ additional memory, which, for fixed $C,$ is $O(\sqrt{N}),$ where
$N = C n^2$ is the number of elements in the matrix.

@@

We want to obtain an upper bound, in machine words, for the value of $\sigma_i + c_i$
in~\eqref{eq:sigmai_and_zi}.
Define $L = (B-1)^2 M.$ By~\eqref{eq:product_inequality} and~\eqref{eq:pi_bound}, we have
$$L < p_1 p_2 < (\mu / 2)^2 = \mu^2 / 4.$$

By~\eqref{eq:ci_bound}, we have $c_i \le L;$ and by~\eqref{eq:sigmai_bound}, we have
$\sigma_i < (B-1)M \le L.$ But then
$$\sigma_i + c_i < 2L < \mu^2 / 2.$$

Note this is far from being a tight bound in numerical sense:
in practice, the value of
$\frac{L}{(B-1)M}=B-1$ is going to be much larger than 1, meaning that the exact upper bound on
$c_i + \sigma_i$ is going to be closer to $\mu^2 / 4$ than to $\mu^2 / 2;$ but this is still a
tight bound in the sense that maximum possible $c_i + \sigma_i$ would not into one machine word.
This bound also has the advantage that it works for any $B > 1.$
